---
title: "Activity: Bird Species"
author: "Guilherme Lima"
date: "6/7/23"
output: html_notebook
---


# Set Up

```{r}
# Load `tidyverse` and `dcData` packages
library(tidyverse)
library(dcData)


# Load the `OrdwayBirds` data set into your RStudio environment from the `dcData` package
data("OrdwayBirds", package = "dcData")



# Show a few commands used to inspect the data set a few different ways
head(OrdwayBirds)
#glimpse(OrdwayBirds)
summary(OrdwayBirds)
#View(OrdwayBirds)
#names(OrdwayBirds)
```

# Activity

When you're finished with this activity, you will have a graph that shows what time of year various species appear at the Katherine Ordway Natural History Study Area in Inver Grove Heights, MN.

### Step 0

Before we begin, the book gives us some instructions to select a few key variables and clean up the date formatting.  Use the commands given in the book, but be sure to study the functions used so you know what is happening.  In this case, we select `SpeciesName`, `Month`, and `Day`. Then a `mutate()` command converts each variable to character, and then converts the resulting "character string" to a number.

```{r}
# Get the data table & clean up dates (see Data Computing eBook)
OrdwayBirds <-
  OrdwayBirds %>%
  select( SpeciesName, Month, Day ) %>%
  mutate( Month = as.numeric(as.character(Month)), 
          Day = as.numeric(as.character(Day)))


# Inspect resulting data
head(OrdwayBirds)
summary(OrdwayBirds)
```



### Step 1




```{r}
# unique species in original data (including mis-spellings)
n_distinct(OrdwayBirds$SpeciesName)


# unique species in the clean list
data("OrdwaySpeciesNames", package = "dcData")
n_distinct(OrdwaySpeciesNames$SpeciesNameCleaned)

```
275 Unique Species in original data

109 Unique Species in the clean list 

### Step 2
  

```{r}
# inspect OrdwaySpeciesNames
head(OrdwaySpeciesNames)
summary(OrdwaySpeciesNames)

# inner join
Corrected <-
  OrdwayBirds %>%
  inner_join( OrdwaySpeciesNames ) %>% 
  select( Species = SpeciesNameCleaned, Month, Day ) %>%
  na.omit()  # cleaned up the missing ones


# inspect the data after the join
head(Corrected)
summary(Corrected)
```

EBook Q1: Which variable(s) was used for matching cases?
SpeciesName was used for matching cases

EBook Q2: What were the variable(s) that will be added?
There were no varibles added, although since OrdwayBirds already contained Month, Day.

### Step 3

We want to identify the top 5-6 bird species sighted at the Katherine Ordway Natural History Study Area.

```{r}
# count sightings by species in descending order
sightingsCount <- Corrected %>% group_by(Species) %>% summarise(count = n()) %>% arrange(desc(count))



# a hint in the text book recommends displaying the top 10 species to choose our threshold
sightingsCount %>% head(10)
sightingsThreshold <- sightingsCount$Species %>% head(6) 


# define a sightings threshold; produce a table with only birds belonging to major species
Majors <- Corrected %>% filter(Species %in% sightingsThreshold)

```


### Step 4

We now want to produce a month-by-month count for each of the major species, and then construct a chart to tell the story of what time of year the various species appear.

```{r}
# prepare glyph-ready data table called `ByMonth`
ByMonth <- Majors %>% group_by(Species, Month) %>% summarise(count = n()) %>% arrange(Month)

```

Now use the ggplot template below to generate the desired graph. (Remark you can also use esquisser or mplot to create a line chart instead of the barchart)



```{r}
# ggplot commands

ByMonth %>% 
  ggplot(aes(x = Month, y = count, color = Species)) + 
  geom_line()

ByMonth %>% ggplot(aes(x=Month,y=count ,fill=Species)) +
  geom_bar(stat='identity',position='stack', width=.9) 

```

According to the graph, we can answer the questions posed:

##### 1. Which species are present year-round?  
American Goldfinch
Black-capped Chickadee
Slate-colored Junco
Tree Swallow

##### 2. Which species are migratory, that is, primarily present in one or two seasons?
Lincoln's Sparrow
Slate-colored Junco


##### 3. What is the peak month for each major species? (bullet list)
*American Goldfinch: October
*Black-capped Chickadee: November
*Field Sparrow: May
*Lincoln's Sparrow: October
*Slate-colored Junco: October
*Tree Swallow: March




##### 4. Which major species are seen in good numbers for at least 6 months of the year?
American Goldfinch
Black-capped Chickadee
Field Sparrow
Slate-colored Junco
Tree Swallow




